Shaw et al. (2025) Supplementary analysis

### 2025-04-17 scripts/Paper-analysis.Rmd
more cleaning up
cleaning up

**Note** This relies on large downloaded files and is not intended to work out-of-the-box. For example, Roary gene presence/absence tables on my computer are stored in `/Users/Liam/Downloads/06_plaspan/`.

@@

```{r, message=FALSE}
# Theoretical expectation for 6-bp palindrome density
theoreticalPalindromeExpectation = function(p){# p is GC content
  return ((p**2 - p + 1/2)**3)
}
d = read.csv('../test_output_window5000_step500.csv', header=F, stringsAsFactors = F)
colnames(d) = c("PTU", "plasmid", "x", "window_size", "palindrome_count", "palindrome_density", "GC_n")

@@

results = NULL
for (PTU in leading.region.PTUs){
  for (k in c(5,6)){
    scores_df = read.csv(paste0('../test_target_leading_lagging/', PTU, '_k', k, '_scores.csv'), header=T)
    n_targets = nrow(scores_df)
    scores_average <- scores_df %>%
  rowwise() %>%
  mutate(score_avg = mean(c_across(starts_with("score")))) %>%
  ungroup() %>%
  select(kmer, score_avg)
    counts_df = read.csv(paste0('../test_target_leading_lagging/', PTU, '_k', k, '_counts.csv'), header=T)
    counts_average <- counts_df %>%
  rowwise() %>%
  mutate(count_avg = mean(c_across(starts_with("count")))) %>%
  ungroup() %>%
  select(kmer, count_avg)
    results = rbind(results, c(PTU, k, n_targets, median(scores_average$score_avg), median(counts_average$count_avg)))
  }
}

@@

```{r}
scores_df = read.csv('../test_6mer_leading_lagging_scores.csv', header=T)
scores_average <- scores_df %>%
  rowwise() %>%
  mutate(score_avg = mean(c_across(starts_with("score")))) %>%
  ungroup() %>%
  select(kmer, score_avg)  
# Order by average score
scores_average$kmer = ordered(scores_average$kmer,
                              levels=scores_average$kmer[order(scores_average$score_avg)])
scores.plot = ggplot(scores_average, aes(kmer, score_avg))+geom_point()+
  coord_flip()

# Same for counts
counts_df = read.csv('../test_6mer_leading_lagging_counts.csv', header=T)
counts_average <- counts_df %>%
  rowwise() %>%
  mutate(count_avg = mean(c_across(starts_with("count")))) %>%
  ungroup() %>%
  select(kmer, count_avg)

@@

### Short palindromes (k=4, 6)

```{r}
# Pangenome statistics for PTUs
#ptu = read.csv('../data/PTU-pangenome-stats.csv')

# Read in RMES comparisons
df_4 = read.csv('~/Downloads/trieste/test_k4.csv')
df_6 = read.csv('~/Downloads/trieste/test_k6.csv')
df_4_filter = df_4[df_4$PTU %in% ptu$PTU[ptu$pangenome.category=="include"],]
df_6_filter = df_6[df_6$PTU %in% ptu$PTU[ptu$pangenome.category=="include"],]
```

Do the statistical test

```{r}
wilcox.test(df_4_filter$discrepancy_selected, df_4_filter$discrepancy_others, paired = TRUE)
wilcox.test(df_6_filter$discrepancy_selected, df_6_filter$discrepancy_others, paired = TRUE)

### All RM targets (not included in main text)

@@

```{r}
# Read in results for RM targets
df_4 = read.csv('../test_one_per_family_within_range_targets_k4.csv')
df_5 = read.csv('../test_one_per_family_within_range_targets_k5.csv')
df_6 = read.csv('../test_one_per_family_within_range_targets_k6.csv')

>>>>

**Note** This relies on large downloaded files and is not intended to work out-of-the-box. For example, plasmid pangenome gene presence/absence tables are available on the associated Zenodo dataset as `plaspan.tar.gz`. This markdown document assumes they have been extracted to `/Users/Liam/Downloads/plaspan` (replace as appropriate).

@@

```{r, message=FALSE}
# Theoretical expectation for 6-bp palindrome density
theoreticalPalindromeExpectation = function(p){# p is GC content
  return ((p**2 - p + 1/2)**3)
}
d = read.csv('../results/output_window5000_step500.csv', header=F, stringsAsFactors = F)
colnames(d) = c("PTU", "plasmid", "x", "window_size", "palindrome_count", "palindrome_density", "GC_n")

@@

results = NULL
for (PTU in leading.region.PTUs){
  for (k in c(5,6)){
    scores_df = read.csv(paste0('../targets_leading_lagging/', PTU, '_k', k, '_scores.csv'), header=T)
    n_targets = nrow(scores_df)
    scores_average <- scores_df %>%
  rowwise() %>%
  mutate(score_avg = mean(c_across(starts_with("score")))) %>%
  ungroup() %>%
  select(kmer, score_avg)
    counts_df = read.csv(paste0('../targets_leading_lagging/', PTU, '_k', k, '_counts.csv'), header=T)
    counts_average <- counts_df %>%
  rowwise() %>%
  mutate(count_avg = mean(c_across(starts_with("count")))) %>%
  ungroup() %>%
  select(kmer, count_avg)
    results = rbind(results, c(PTU, k, n_targets, median(scores_average$score_avg), median(counts_average$count_avg)))
  }
}

@@

```{r}
scores_df = read.csv('../results/6mer_leading_lagging_scores.csv', header=T)
scores_average <- scores_df %>%
  rowwise() %>%
  mutate(score_avg = mean(c_across(starts_with("score")))) %>%
  ungroup() %>%
  select(kmer, score_avg)  
# Order by average score
scores_average$kmer = ordered(scores_average$kmer,
                              levels=scores_average$kmer[order(scores_average$score_avg)])
scores.plot = ggplot(scores_average, aes(kmer, score_avg))+geom_point()+
  coord_flip()

# Same for counts
counts_df = read.csv('../6mer_leading_lagging_counts.csv', header=T)
counts_average <- counts_df %>%
  rowwise() %>%
  mutate(count_avg = mean(c_across(starts_with("count")))) %>%
  ungroup() %>%
  select(kmer, count_avg)

@@

### All RM targets

@@

```{r}
# Read in results for RM targets
df_4 = read.csv('../results/one_per_family_within_range_targets_k4.csv')
df_5 = read.csv('../results/one_per_family_within_range_targets_k5.csv')
df_6 = read.csv('../results/one_per_family_within_range_targets_k6.csv')
